\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-29 k-planar.tex
minor updates
comments on map graph proof

\begin{abstract}
Dujmovi\'c~et~al.~[\emph{J.~ACM}~'20] recently proved that every planar graph is isomorphic to a subgraph of the strong product of a bounded treewidth graph and a path. Analogous results were obtained for graphs of bounded Euler genus or apex-minor-free graphs. These tools have been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and adjacency labelling. This paper proves analogous product structure theorems for various non-minor-closed classes. One noteable example is $k$-planar graphs (those with a drawing in the plane in which each edge is involved in at most $k$ crossings). We prove that every $k$-planar graph is isomorphic to a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. This leads to analogous results for certain types of map graphs, string graphs, graph powers, and nearest neighbour graphs.
\end{abstract}
\end{titlepage}
\pagenumbering{roman}
\tableofcontents
\newpage

@@

\begin{table}[h]
  \centering{
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $4^*$ & $2$ & \cite{bose.morin.ea:optimal} \\
      planar & $6^*$ & $1$ & \cite{UWY} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
      genus $g$ & $2g+6$ & 1 & \cite{UWY} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-nearest-neighbour & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \multicolumn{4}{l}{$^*$ these bounds also apply to the simple treewidth of $H$.}
    \end{tabular}
  }
  \caption{Known product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

@@

\begin{lem}
\label{MewMapGraphLemma}
For every %all
integer %s $g\geq 0$ and
 $d\geq 3$, every $d$-map graph is a spanning subgraph of a $d$-framed graph.
\end{lem}

\note{PM}{Suggested new Proof}
\begin{proof}
  Let $G_0$ be a graph embedded in the plane, with each face labelled a nation or a face, and where each vertex of $G_0$ is incident with at most $d$ nations and let $G$ be the corresponding map graph.  Let $G_0^*$ be the dual graph of $G_0$, so that each vertex of $G_0^*$ corresponds to face of $G_0$ and vice-versa.  Thus, each vertex of $G_0^*$ is a nation vertex or a lake vertex.

  Let $x$ be a vertex of $G_0$, let $F_x$ be the corresponding face of $G_0^*$, and let $v_0,\ldots,v_r$ be the facial cycle of $F_x$.  Let $C:=w_0,\ldots,w_s$ be the subsequence of $v_0,\ldots,v_r$ consisting of only the nation vertices.  Since $x$ is incident to at most $d$ nations, $r\le d$.  We call $C_x$ the \emph{nation cycle} of $F_x$.

  Let $G_1$ be a plane supergraph of $G_0^*$ obtained by adding each nation cycle of each face $F$ of $G^0$ and then triangulating each face with more than $d$ vertices on its boundary.  Each nation cycle is now a face of $G_1$ and no face of $G_1$ has more than $d$ vertices on its boundary. Let $\widehat{G}_1$ be the $d$-framed graph whose frame is $G_1$.

  Since $\widehat{G}_1$ is a $d$-framed graph, it suffices to show that these definitions imply that $E(G)\subseteq E(\widehat{G}_1)$.  Indeed, if $vw\in E(G)$ then the nation faces corresponding to $v$ and $w$ have a common vertex $x$ on their boundary. The vertex $x$ corresponds to a face $F_x$ in $G_0^*$ and the facial cycle of $F_x$ contains $v$ and $w$.  Therefore, the nation cycle $C_x$ of $F_x$ contains $v$ and $w$. Since $C_x$ a face in $G_1$ the vertices $vw\in E(G_1)$.
\end{proof}

>>>>

\begin{abstract}
Dujmovi\'c~et~al.~[\emph{J.~ACM}~'20] recently proved that every planar graph is isomorphic to a subgraph of the strong product of a bounded treewidth graph and a path. Analogous results were obtained for graphs of bounded Euler genus or apex-minor-free graphs. These tools have been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and adjacency labelling. This paper proves analogous product structure theorems for various non-minor-closed classes. One noteable example is $k$-planar graphs (those with a drawing in the plane in which each edge is involved in at most $k$ crossings). We prove that every $k$-planar graph is isomorphic to a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a more general setting based on so-called shortcut systems, which are of independent interest. This leads to analogous results for certain types of map graphs, string graphs, graph powers, and nearest neighbour graphs.
\end{abstract}
\end{titlepage}
\pagenumbering{roman}
\tableofcontents
\newpage

@@

\begin{table}[h]
  \centering{
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $4^*$ & $2$ & \cite{bose.morin.ea:optimal} \\
      planar & $6^*$ & $1$ & \cite{UWY} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
      genus $g$ & $2g+6$ & 1 & \cite{UWY} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-nearest-neighbour & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \multicolumn{4}{l}{$^*$ these bounds also apply to the simple treewidth of $H$.}
    \end{tabular}
  }
  \caption{Product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

@@

\begin{lem}
\label{MewMapGraphLemma}
For every %all
integer %s $g\geq 0$ and
 $d\geq 3$, every $d$-map graph is a spanning subgraph of a $d$-framed graph.
\note{DW}{Does this need to be ``spanning subgraph of a $d$-framed multigraph''?}
\end{lem}

\note{PM}{Suggested new Proof}
\note{DW}{I am happy with the new proof, once we address the issues.}

\begin{proof}
  Let $G_0$ be a graph embedded in the plane, with each face labelled a nation or a face, and where each vertex of $G_0$ is incident with at most $d$ nations and let $G$ be the corresponding map graph.  Let $G_0^*$ be the dual graph of $G_0$. So the vertices of $G_0^*$ correspond to faces of $G_0$, and each vertex of $G_0^*$ is a nation vertex or a lake vertex.

  Let $x$ be a vertex of $G_0$, let $F_x$ be the corresponding face of $G_0^*$, and let $v_0,\ldots,v_s$ be the facial cycle of $F_x$.  Let $C:=w_0,\ldots,w_r$ be the subsequence of $v_0,\ldots,v_s$ consisting of only the nation vertices.  Since $x$ is incident to at most $d$ nations, $r\le d$ \note{DW}{Why not $r+1\leq d$? Do we mean that $v_0=v_r$? I suggest we write $(v_1,\dots,v_s)$ and $(w_1,\dots,w_r)$. }.  Call $C_x$ the \defin{nation cycle} of $F_x$. \note{DW}{What if there are 1 or 2 nations incident to $x$? Is $C_x$ a cycle? Do we need parallel edges here?}

  Let $G_1$ be a plane supergraph of $G_0^*$ obtained by adding each nation cycle of each face $F$ of $G^0$ \note{DW}{We mean `` face $F$ of $G^*_0$'', right?} and then triangulating each face with more than $d$ vertices on its boundary.  Each nation cycle is now a face of $G_1$ and no face of $G_1$ has more than $d$ vertices on its boundary. Let $\widehat{G}_1$ be the $d$-framed graph whose frame is $G_1$.

By definition, $V(G)=V(\widehat{G}_1)$. To prove the claim, it suffices to show that $E(G)\subseteq E(\widehat{G}_1)$.  Indeed, if $vw\in E(G)$ then the nation faces corresponding to $v$ and $w$ have a common vertex $x$ on their boundary. The vertex $x$ corresponds to a face $F_x$ in $G_0^*$ and the facial cycle of $F_x$ contains $v$ and $w$.  Therefore, the nation cycle $C_x$ of $F_x$ contains $v$ and $w$. Since $C_x$ bounds a face in $G_1$, $vw\in E(\widehat{G}_1)$.
\end{proof}
